---
title: "DS2020 Final Project Team 29"
author: "Laura Arazan"
date: "2025-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Customer Personality Analysis


### Introduction

This dataset is a collection of information on a company's customers. I found it on kaggle distributed by Dr. Omar Romero-Hernandez. I am interested in discovering how to best find a specific set of information (in this case customers) that best fits a set of requirements (in this case the most beneficial customers to a company). While testing on this dataset I will be able to narrow what the most useful tools to me are and gaining familiarity using them. I think this dataset is similar to what a company could give me in the future with the same goal of finding the best candidates to focus their attention on.

I will be reading the data in, cleaning it, looking for points of interest and familiarizing myself with the format and data, exploring different relationships between data points, comparing models, and narrowing down candidates until I find a set of data points that best fit what a company's ideal customer is composed of.

The final result could be used for a marketing or ad campaign targeting the most effective group of customers in this company.

The data has a couple points of concern that I noticed; there isn't a range on when the data was collected or how it was collected, one of the options for Marital_Status is called "absurd", there isn't any data on customers with a maximum education below a bachelor's degree, no location(s) are given, and there is no explanation given for what types of discounts customers used or what type of campaign(s) they accepted.


### Data

Since the data is in a csv file on kaggle we can download it and read it in using the readr library in r.

```{r}
library(tidyverse)
library(readr)
library(ggplot2)

```


```{r}
df <- read_delim("marketing_campaign.csv", delim = "\t")

summary(df)
head(df)
str(df)

```
Condense variables age, Total_Children, Total_Spent, and Total_Campaign_Accepted

```{r}
df$Age <- 2025 - df$Year_Birth
df$Total_Children <- df$Kidhome + df$Teenhome
df$Total_Spent <- rowSums(df[, c("MntWines", "MntFruits", "MntMeatProducts", 
                                 "MntFishProducts", "MntSweetProducts", "MntGoldProds")])
df$Total_Campaign_Accepted <- rowSums(df[, c("AcceptedCmp1", "AcceptedCmp2", 
                                             "AcceptedCmp3", "AcceptedCmp4", "AcceptedCmp5")])
summary(df$Income)

```

Clean the data

```{r}
missing <- colSums(is.na(df))
missing[missing > 0]

df <- df %>% drop_na(Income)
missing <- colSums(is.na(df))
missing[missing > 0]

```

Explore relations

```{r}
ggplot(df, aes(x = Income, y = Total_Spent)) +
  geom_point(alpha = 0.4, color = "lightblue") +
  geom_smooth(method = "lm", color = "pink", se = FALSE) +
  labs(title = "Income vs Total Spending", x = "Income ($)", y = "Total Spend ($)")

ggplot(df, aes(x = Age, y = Total_Spent)) +
  geom_point(alpha = 0.4, color = "lightgreen") +
  geom_smooth(method = "loess", color = "pink", se = FALSE) +
  labs(title = "Age vs Total Spending", x = "Age", y = "Total Spend ($)")

ggplot(df, aes(x = Total_Campaign_Accepted)) +
  geom_bar(fill = "purple") +
  labs(title = "Number of Campaigns Accepted", x = "Campaigns Accepted", y = "Customer Count")
```

Discovering the top 20% of customers

```{r}
top_spenders <- df %>% 
  filter(Total_Spent > quantile(Total_Spent, 0.80))

ggplot(top_spenders, aes(x = Marital_Status, fill = Education)) +
  geom_bar(position = "dodge") +
  labs(title = "Top Spenders by Marital Status & Education", x = "Marital Status")

```

What does an ideal customer look like?

```{r, fig.height = 12}
ideal <- df %>%
  filter(
    Total_Spent > quantile(Total_Spent, 0.80),
    Total_Campaign_Accepted > 0,
    Age >= 30, Age <= 60,
    Income >= median(Income, na.rm=TRUE)
  )

summary(ideal)

df_seperated_customers <- df %>%
  mutate(
    Ideal_Customer = ifelse(
      Total_Spent > quantile(Total_Spent, 0.80, na.rm = TRUE) &
      Total_Campaign_Accepted > 0 &
      Age >= 30 & Age <= 60 &
      Income >= median(Income, na.rm = TRUE),
      "Ideal", "Other"
    )
  )

# Select numeric columns (excluding ID)
numeric_cols <- df_seperated_customers %>%
  select(where(is.numeric)) %>%
  select(-ID) %>%
  colnames()

# Reshape to long format
df_long <- df_seperated_customers %>%
  select(all_of(numeric_cols), Ideal_Customer) %>%
  pivot_longer(
    cols = -Ideal_Customer,
    names_to = "Variable",
    values_to = "Value"
  )

# Plot faceted boxplots
ggplot(df_long, aes(x = Ideal_Customer, y = Value, fill = Ideal_Customer)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  facet_wrap(~ Variable, scales = "free", ncol = 3) +
  labs(title = "Comparison of All Numeric Variables by Customer Type",
       x = "Customer Type", y = "Value") +
  scale_fill_manual(values = c("Ideal" = "orange", "Other" = "gray")) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

Idea customers are most likely

- Aged 30–60

- Spend a lot across product categories

- Responsive to campaigns

- Mid-to-high income

- Educated and likely married
