---
title: "DS2020 Final Project Team 29"
author: "Laura Arazan"
date: "2025-05-12"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Customer Personality Analysis


### Introduction

This dataset is a collection of information on a company's customers. I found it on kaggle distributed by Dr. Omar Romero-Hernandez. I am interested in discovering how to best find a specific set of information (in this case customers) that best fits a set of requirements (in this case the most beneficial customers to a company). While testing on this dataset I will be able to narrow what the most useful tools to me are and gaining familiarity using them. I think this dataset is similar to what a company could give me in the future with the same goal of finding the best candidates to focus their attention on.

I will be reading the data in, cleaning it, looking for points of interest and familiarizing myself with the format and data, exploring different relationships between data points, comparing models, and narrowing down candidates until I find a set of data points that best fit what a company's ideal customer is composed of.

The final result could be used for a marketing or ad campaign targeting the most effective group of customers in this company.

The data has a couple points of concern that I noticed; there isn't a range on when the data was collected or how it was collected, one of the options for Marital_Status is called "absurd", there isn't any data on customers with a maximum education below a bachelor's degree, no location(s) are given, and there is no explanation given for what types of discounts customers used or what type of campaign(s) they accepted.


### Data

Since the data is in a csv file on kaggle I downloaded it and read it in using the readr library in R.

```{r}
library(tidyverse)
library(readr)
library(ggplot2)

```
Then, I read it into R and ran the dataset through some functions to get a better understanding of it.

```{r}
df <- read_delim("marketing_campaign.csv", delim = "\t")

summary(df)
head(df)
str(df)

```
Here I condensed variables into; Age, Total_Children, Total_Spent, and Total_Campaign_Accepted.

```{r}
df$Age <- 2025 - df$Year_Birth
df$Total_Children <- df$Kidhome + df$Teenhome
df$Total_Spent <- rowSums(df[, c("MntWines", "MntFruits", "MntMeatProducts", 
                                 "MntFishProducts", "MntSweetProducts", "MntGoldProds")])
df$Total_Campaign_Accepted <- rowSums(df[, c("AcceptedCmp1", "AcceptedCmp2", 
                                             "AcceptedCmp3", "AcceptedCmp4", "AcceptedCmp5")])

```

Afterwards, I cleaned the data by checking for any missing or null values. Seeing income only had 24 missing values out of 2240 data points, I proceeded to cut the missing rows out of the dataset.

```{r}
missing <- colSums(is.na(df))
missing[missing > 0]
summary(df$Income)

df <- df %>% drop_na(Income)
missing <- colSums(is.na(df))
missing[missing > 0]

```

Then I explored the relationships between categories that stood out to me, including; income vs total spent, age vs total spent, and how many customers accepted this company's campaigns.

```{r}
ggplot(df, aes(x = Income, y = Total_Spent)) +
  geom_point(alpha = 0.4, color = "lightblue") +
  geom_smooth(method = "lm", color = "pink", se = FALSE) +
  labs(title = "Income vs Total Spending", x = "Income ($)", y = "Total Spend ($)")

ggplot(df, aes(x = Age, y = Total_Spent)) +
  geom_point(alpha = 0.4, color = "lightgreen") +
  geom_smooth(method = "loess", color = "pink", se = FALSE) +
  labs(title = "Age vs Total Spending", x = "Age", y = "Total Spend ($)")

ggplot(df, aes(x = Total_Campaign_Accepted)) +
  geom_bar(fill = "purple") +
  labs(title = "Number of Campaigns Accepted", x = "Campaigns Accepted", y = "Customer Count")
```

I discovered the top 20% of customers by filtering customers that were in the top 20% of money spent at this company. Then, I made a plot of the top spending customers and compared their marital status vs their education. I saw that the top spenders tend to be married or have a bachelor's degree, and this dataset seems to exclude any customers with an education below a bachelor’s degree. Also, As income increases so does the amount of money spent, there seems to be an increase in money spent around the middle of the customers’ income, customers tend to start purchasing from this company around 30 years old and begin to drop off around 60 years old, and most customers don’t participate in this company’s campaigns.



```{r}
top_spenders <- df %>% 
  filter(Total_Spent > quantile(Total_Spent, 0.80))

ggplot(top_spenders, aes(x = Marital_Status, fill = Education)) +
  geom_bar(position = "dodge") +
  labs(title = "Top Spenders by Marital Status & Education", x = "Marital Status")

```

What does an ideal customer look like? Taking into account what the graphs above, I filtered the top 20% of spending customers, who accepted at least one campaign, were from 30 to 60 (inclusive), and had a median income at or above that of all of the customers sampled.

```{r, fig.height = 12}
ideal <- df %>%
  filter(
    Total_Spent > quantile(Total_Spent, 0.80),
    Total_Campaign_Accepted > 0,
    Age >= 30, Age <= 60,
    Income >= median(Income, na.rm=TRUE)
  )

summary(ideal)

df_seperated_customers <- df %>%
  mutate(
    Ideal_Customer = ifelse(
      Total_Spent > quantile(Total_Spent, 0.80, na.rm = TRUE) &
      Total_Campaign_Accepted > 0 &
      Age >= 30 & Age <= 60 &
      Income >= median(Income, na.rm = TRUE),
      "Ideal", "Other"
    )
  )

# Select numeric columns (excluding ID)
numeric_cols <- df_seperated_customers %>%
  select(where(is.numeric)) %>%
  select(-ID) %>%
  colnames()

# Reshape to long format
df_long <- df_seperated_customers %>%
  select(all_of(numeric_cols), Ideal_Customer) %>%
  pivot_longer(
    cols = -Ideal_Customer,
    names_to = "Variable",
    values_to = "Value"
  )

# Plot faceted boxplots
ggplot(df_long, aes(x = Ideal_Customer, y = Value, fill = Ideal_Customer)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  facet_wrap(~ Variable, scales = "free", ncol = 3) +
  labs(title = "Comparison of All Numeric Variables by Customer Type",
       x = "Customer Type", y = "Value") +
  scale_fill_manual(values = c("Ideal" = "orange", "Other" = "gray")) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

Idea customers are most likely

- Aged 30–60

- Spend a lot across product categories

- Responsive to campaigns

- Mid-to-high income

- Educated and likely married
